---
title: "Health Data Final Project"
author: "R2"
date: "2023-02-16"
output: html_document
---

```{r}

library(readxl)
library(tidyverse)
library(dplyr)
library(tidyselect)
library(lubridate)
library(ggplot2)

```

Load Data Sets:
```{r}
med19= read_excel("Data/Medical 2019.xlsx") %>% 
  janitor:: clean_names()
med20= read_excel("Data/Medical 2020.xlsx") %>% 
  janitor:: clean_names()
rx20= read_excel("Data/Rx 2020.xlsx") %>% 
  janitor:: clean_names()
rx19= read_excel("Data/Rx 2019.xlsx") %>% 
  janitor:: clean_names()
risk20= read_excel("Data/Risk 2020.xlsx") %>% 
  janitor:: clean_names()
risk19=read_excel("Data/Risk 2019.xlsx") %>% 
  janitor:: clean_names()
biom19=read_excel("Data/Biometrics 2019.xlsx") %>% 
  janitor:: clean_names()
biom20=read_excel("Data/Biometricsx 2020.xlsx") %>% 
  janitor:: clean_names()
```

```{r}
#remove NA rows if data is missing bps values
biom19 = biom19 %>% drop_na(bps)
biom20 = biom20 %>% drop_na(bps)

#createhypertensive group
htdrug19 = left_join(biom19, rx19, by = "person")
htdrug19 = htdrug19 %>% 
  rename(bps19 = bps) %>% 
  filter(bps19 >= 130) %>% 
  select(person, from, units, cost, brand, class, drug, gender.x, age_range.x, glucose, bmi, bpd, bps19, hdl, ldl, chol)

counter_19 = htdrug19 %>% 
  group_by(person) %>% 
  count(person)

htdrug20 = left_join(biom20, rx20, by = "person")
htdrug20 = htdrug20 %>% 
  rename(bps20 = bps) %>% 
  select(person, from, units, cost, brand, class, drug, gender.x, age_range.x, glucose, bmi, bpd, bps20, hdl, ldl, chol)

htdrug = merge(htdrug19, htdrug20, by = 'person')
```
```{r}
```{r}
#remove NA rows if data is missing bps values
biom19 = biom19 %>% drop_na(bps)
biom19 = biom19 %>% rename(bps19 = bps)
biom20 = biom20 %>% drop_na(bps)
rx19 = rx19 %>% drop_na(class)
rx20 = rx20 %>% drop_na(class)
```

```{r}
druguse19 = rx19 %>% 
  group_by(person) %>% 
  mutate(htuse19 = case_when(
    (any(class == "CALCIUM CHANNEL BLOCKERS" | class == "ANTIHYPERTENSIVES" | class == "BETA BLOCKERS"))  ~ "yes", 
  TRUE ~ "no")) %>% 
  distinct(person, htuse19) 

druguse19 %>% group_by(htuse19) %>% count()
  
druguse20 = rx20 %>% 
  group_by(person) %>% 
  mutate(htuse20 = case_when(
    (any(class == "CALCIUM CHANNEL BLOCKERS" | class == "ANTIHYPERTENSIVES" | class == "BETA BLOCKERS"))  ~ "yes", 
  TRUE ~ "no")) %>%
   distinct(person, htuse20)

druguse20 %>% group_by(htuse20) %>% count()
```


```{r create new set}
druguse = merge(druguse19, druguse20, "person")

druguse = druguse %>% 
  group_by(person) %>% 
  mutate(htgroup = case_when(
    htuse19 == "yes" & htuse20 == "yes" ~ "consistent",
    htuse19 == "yes" & htuse20 == "no" ~ "stopped2019",
    htuse19 == "no" & htuse20 == "yes" ~ "started2020",
    TRUE ~ "never started"
  )) %>% 
  select(person, htgroup, htuse19, htuse20)

checkdrug = druguse %>% 
  distinct(person, htgroup) %>% 
  group_by(htgroup) %>% 
  count(htgroup)

```

```{r}
#class of hypertensives 
biom = inner_join(biom19, biom20, by = "person")
biom = biom %>% drop_na(bps) %>% rename(bps20 = bps) 

#hypertensive filter
ht = biom %>% filter(bps19 >= 130)

htdrug = merge(x = ht, y= druguse, by = "person", all.x=TRUE)

htdrug = htdrug %>% 
  mutate_at(vars(htgroup), ~replace_na(., "never started"))

htdrug = htdrug %>%
  distinct()

trend = htdrug %>% 
  distinct(person, htgroup) %>% 
  group_by(htgroup) %>% 
  count(htgroup)

```

```{r outcomes}
#outcomes? did bp decrease and by how much 

htdrug$bpsdiff = as.numeric(htdrug$bps20) - as.numeric(htdrug$bps19)

bpscalculation = htdrug %>% 
  distinct(bps20, bps19, bpsdiff, htgroup, person) %>%
  group_by(htgroup) %>% 
  mutate(averagediff = mean(bpsdiff))
  
bpsoutcome = bpscalculation %>% 
  group_by(htgroup) %>% 
  distinct(averagediff, htgroup)
```

```{r pharmaceutical cost}
#how much more do hypertensives cost (general pharmaceuticals)
rxcombined = rbind(rx19, rx20)
htrxcost = merge(htdrug, rxcombined, by = "person")

drugdiff = htrxcost %>% 
  group_by(person) %>% 
  mutate(total_cost = sum(cost)) 

drugdiff1 = drugdiff %>% 
  distinct(total_cost, htgroup, person) %>% 
  group_by(htgroup) %>% 
  mutate(groupavgcost = mean(total_cost)) %>% 
  distinct(htgroup, groupavgcost)

drugdiffht = htrxcost %>% 
  filter(!str_detect(person, '311-01')) %>%
  group_by(person) %>% 
  filter(
     class == "CALCIUM CHANNEL BLOCKERS" | class == "ANTIHYPERTENSIVES" | class == "BETA BLOCKERS"
  ) 

drugdiffht = drugdiffht %>% 
  mutate(htcost = sum(cost)) %>% 
  group_by(htgroup) %>% 
  mutate(grouphtcost = mean(htcost)) %>% 
  distinct(htgroup, grouphtcost)
 
```

```{r ER }
#how many ER visits a person had in 2019, then 2020 
#293 
er19 = med19 %>% 
  group_by(person) %>% 
  mutate(iner19 = case_when(
    (any(er==1))  ~ "yes", TRUE ~ "no" )) %>%
  distinct(iner19, person) 

er19 %>% 
  group_by(iner) %>% 
  count()

er20 = med20 %>% 
  group_by(person) %>% 
  mutate(iner20 = case_when(
    (any(er==1))  ~ "yes", TRUE ~ "no" )) %>%
  distinct(iner20, person) #393 visits 

ermed = merge(er19, er20 by = "person")
#repeat hospitalization 

htdrug19er = merge(x = htdrug, y= er19, by = "person", all.x=TRUE)
htdrug19er %>% 
  group_by(htgroup, iner19) %>% 
  count()
htdrug20er = merge(x = htdrug, y= er20, by = "person", all.x=TRUE)
htdrug20er %>% 
  group_by(htgroup, iner20) %>% 
  count()
#merge ER visits with druguse category 
```

```{r medical cost}
medcombined = rbind(med19, med20)
htmedcost = merge(htdrug, medcombined, by = "person")

#generalcost of care 
personmedcosts = htmedcost %>% 
  group_by(person) %>% 
  mutate(total_cost = sum(cost)) 

drugmeddiff = personmedcosts %>% 
  distinct(total_cost, htgroup, person) %>% 
  group_by(htgroup) %>% 
  mutate(groupavgcost = mean(total_cost)) %>% 
  distinct(htgroup, groupavgcost)

#cost of HT related care
drugdiffht = htmedcost %>% 
  filter(!str_detect(person, '311-01')) %>%
  group_by(person) %>% 
  filter(grepl("^N18|^I25|^I10|^I25|^I21", icd10) | grepl("^N18|^I25|^I10|^I25|^I21", icd2)) %>%
  mutate(htcost = sum(cost))

drugdiffht %>% 
  group_by(htgroup) %>% 
  mutate(htmedcost = mean(htcost)) %>% 
  distinct(htgroup, htmedcost)

```



